Uncertainty propagation in deep neural networks for sparse coding

### 2018-03-10 paper/BayesianLISTA.tex

The paper is organised as following: first, we review neural networks for sparse coding in section \ref{sec:nn_sc} and present a novel Bayesian neural network in section \ref{sec:bayesian_lista}. We then describe uncertainty propagation to obtain the spike-and-slab distribution for an output of a network in section \ref{sec:fprop} and how probabilistic backpropagation is organised to update distributions of weights in section \ref{sec:backpropagation}. After that, we show the experimental results of the algorithm in section \ref{sec:experiments} and finish with discussion and conclusions in sections \ref{sec:discussion} and \ref{sec:conclusions}.

\section{\uppercase{Neural networks for sparse coding}}
\label{sec:nn_sc}
The sparse coding and sparse inverse problems are the linear regression with the assumption of sparse weights
\begin{equation}
\label{eq:regression_problem}
\mathbf{y} = \mathbf{X}\boldsymbol\beta + \varepsilon
\end{equation}
where $\mathbf{y} \in \mathbb{R}^K$ is the observation, $\mathbf{X} \in \mathbb{R}^{K \times D}$ is the design matrix, $\boldsymbol\beta \in \mathbb{R}^D$ is the unknown vector of weight with many elements $\beta_d$ equal to zero. $\varepsilon \in \mathbb{R}^K$ is the independent Gaussian noise.

@@

\section{\uppercase{Experiments}}
\label{sec:experiments}
The proposed Bayesian \textsc{lista} is evaluated in the context of the sparse inverse problem, where the number of measurements $K$ is much smaller than the dimensionality of the vector $\boldsymbol\beta$.

We compare the proposed Bayesian \textsc{lista} with the classical \textsc{lista}~\citep{gregor2010learning} in terms of the predictive accuracy. As baselines we also use \textsc{ista} \citep{daubechies2004iterative} and Fast \textsc{ista} (\textsc{fista}) \citep{beck2009fast}. \textsc{fista} adds the momentum to \textsc{ista} and improves its convergence speed. We limit the number of iterations in these algorithms with $L$ that is the same as the number of layers in Bayesian \textsc{lista} and \textsc{lista}. To measure the performance of the algorithms the normalised mean square error(\textsc{nmse}) and F measure are used. \textsc{nmse} for a batch of data $\{\boldsymbol\beta^{(n)}\}_{n=1}^{N}$ and estimators $\{\widehat{\boldsymbol\beta}^{(n)}\}_{n=1}^{N}$ is computed as
\begin{equation}
\text{\textsc{nmse}} = \frac{1}{N}\sum\limits_{n=1}^N\sqrt{\frac{\sum\limits_{d=1}^D\left(\widehat{\beta}_{d}^{(n)} - \beta_d^{(n)}\right)^2}{\sum\limits_{d=1}^D\left(\beta_{d}^{(n)}\right)^2}}
\end{equation}
In sparse coding it is usually more important to obtain the correct locations of spikes (i.e zeros) and slabs (i.e. non-zeros) in the estimator than to minimise \textsc{nmse}. The problem is therefore viewed as a skewed two-class classification problem where the number of spikes is higher than the number of slabs. F measure is used to evaluate the accuracy of such problems. 
F measure is computed as the harmonic mean of precision and recall
\begin{equation}
\text{F measure} = 2\dfrac{\text{precision}\cdot\text{recall}}{\text{precision} + \text{recall}}
\end{equation} 
where precision is the fraction of estimated slab locations that are correct, recall is the fraction of true slab locations among all predicted slab locations.
We demonstrate the performance on small datasets to highlight that the proposed algorithm can infer accurate predictions when the dataset size is not sufficient for \textsc{lista} to learn anything.

\subsection{\uppercase{Predictive performance on synthetic data}}
First, the predictive performance of the proposed Bayesian \textsc{lista} is analysed on synthetic data. We generate $N_\text{train}=500$ sparse coefficients vectors $\boldsymbol\beta_n$ each of size $D = 100$. Coefficients $\boldsymbol\beta$ are generated from the spike-and-slab distribution with truncated slab: each component of $\beta_{nd}$ is zero with probability $0.8$ or is from standard Gaussian distribution without interval $(-0.1, 0.1)$ with probability $0.2$. Truncation of the zeros allows to remove from support pattern values with insignificant contributions \citep{xin2016maximal}. To simulate sparse observations, we have generated the random Gaussian design matrix $\mathbf{X} \in \mathbb{R}^{K \times D}$.  The observations are generated according to the \ref{eq:regression_problem} with zero-mean Gaussian noise with standard deviation $0.5$. $\lambda$ is assumed fixed as $0.1$.

In the Figure \ref{fig:number_of_layers_synthetic} prediction performance for different number of layers $L$ is presented. The observation size is set as $K=50$.

In the Figure \ref{fig:unsersampling_synthetic} prediction performance for different observation sizes $K$ is presented. The number of layers is set as $L=4$.

>>>>

The paper is organised as following: first, we review neural networks for sparse coding in section \ref{sec:nn_sc} and present a novel Bayesian neural network in section \ref{sec:bayesian_lista}. We then describe uncertainty propagation to obtain the spike and slab distribution for an output of a network in section \ref{sec:fprop} and how probabilistic backpropagation is organised to update distributions of weights in section \ref{sec:backpropagation}. After that, we show the experimental results of the algorithm in section \ref{sec:experiments} and finish with discussion and conclusions in sections \ref{sec:discussion} and \ref{sec:conclusions}.

\section{\uppercase{Neural networks for sparse coding}}
\label{sec:nn_sc}
The sparse coding problem is the linear regression with the assumption of sparse weights
\begin{equation}
\label{eq:regression_problem}
\mathbf{y} = \mathbf{X}\boldsymbol\beta + \varepsilon
\end{equation}
where $\mathbf{y} \in \mathbb{R}^K$ is the observation, $\mathbf{X} \in \mathbb{R}^{K \times D}$ is the design matrix, $\boldsymbol\beta \in \mathbb{R}^D$ is the unknown vector of weight with many elements $\beta_d$ equal to zero. $\varepsilon \in \mathbb{R}^K$ is the independent Gaussian noise.

@@

\section{\uppercase{Experiments}}
\label{sec:experiments}
The proposed Bayesian \textsc{lista} is evaluated in the context of the sparse coding problem with underdefined dictionary, where the number of measurements $K$ is much smaller than the dimensionality of the vector $\boldsymbol\beta$.

We compare the proposed Bayesian \textsc{lista} with the classical \textsc{lista}~\citep{gregor2010learning} in terms of the predictive accuracy. As baselines we also use \textsc{ista} \citep{daubechies2004iterative} and Fast \textsc{ista} (\textsc{fista}) \citep{beck2009fast}. \textsc{fista} adds the momentum to \textsc{ista} and improves its convergence speed. We limit the number of iterations in these algorithms with $L$ that is the same as the number of layers in Bayesian \textsc{lista} and \textsc{lista}. To measure the performance of the algorithms the normalised mean square error(\textsc{nmse}) and F measure are used. \textsc{nmse} for a batch of data $\{\boldsymbol\beta^{(n)}\}_{n=1}^{N}$ and estimators $\{\widehat{\boldsymbol\beta}^{(n)}\}_{n=1}^{N}$ is computed as
\begin{equation}
\text{\textsc{nmse}} = \frac{1}{N}\sum\limits_{n=1}^N\sqrt{\frac{\sum\limits_{d=1}^D\left(\widehat{\beta}_{d}^{(n)} - \beta_d^{(n)}\right)^2}{\sum\limits_{d=1}^D\left(\beta_{d}^{(n)}\right)^2}}
\end{equation}
In sparse coding it is usually more important to obtain the correct locations of spikes (i.e zeros) and slabs (i.e. non-zeros) in the estimator than to minimise \textsc{nmse}. The problem is therefore viewed as a skewed two-class classification problem where the number of spikes is higher than the number of slabs. F measure is used to evaluate the accuracy of such problems. 
F measure is computed as the harmonic mean of precision and recall
\begin{equation}
\text{F measure} = 2\dfrac{\text{precision}\cdot\text{recall}}{\text{precision} + \text{recall}}
\end{equation} 
where precision is the fraction of estimated slab locations that are correct, recall is the fraction of true slab locations among all predicted slab locations.
We demonstrate the performance on small datasets to highlight that the proposed algorithm can infer accurate predictions when the dataset size is not sufficient for \textsc{lista} to learn.

\subsection{\uppercase{Predictive performance on synthetic data}}
First, the predictive performance of the proposed Bayesian \textsc{lista} is analysed on synthetic data. We generate $N_\text{train}=500$ sparse coefficients vectors $\boldsymbol\beta_n$ each of size $D = 100$. Coefficients $\boldsymbol\beta$ are generated from the spike and slab distribution with truncated slab: each component of $\beta_{nd}$ is zero with probability $0.8$ or is from standard Gaussian distribution without interval $(-0.1, 0.1)$ with probability $0.2$. Truncation of the zeros allows to remove from support pattern values with insignificant contributions \citep{xin2016maximal}. To simulate sparse observations, we have generated the random Gaussian design matrix $\mathbf{X} \in \mathbb{R}^{K \times D}$.  The observations are generated according to the \ref{eq:regression_problem} with zero-mean Gaussian noise with standard deviation $0.5$. $\lambda$ is assumed fixed as $0.1$.

In the Figure \ref{fig:number_of_layers_synthetic} prediction performance for different number of layers $L$ is presented. The observation size is set as $K=50$. In both experiments the performance of the Bayesian \textsc{lista} is better than the performance of \textsc{lista}. Although the baseline \textsc{ista} and \textsc{fista} show better performance in terms of F measure, Bayesian \textsc{lista} has lower \textsc{nmse}.

In the Figure \ref{fig:unsersampling_synthetic} prediction performance for different observation sizes $K$ is presented. The number of layers is set as $L=4$ which is the value of the poorest performance of Bayesian \textsc{lista} according to the previous experiment. Bayesian \textsc{lista} provides better results than {lista} for $K \ge 30$ and better results than other algorithms for $K \ge 70$.
